---
title: "GCPS Color Style Guide"
subtitle: "Base palette, scales, and data visualization use cases"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Core palette --------------------------------------------------------------

gcps_base <- c(
  maroon = "#660000",
  blue   = "#2F5FB3",
  teal   = "#007C91",
  green  = "#5E8C31",
  violet = "#6A4CC3",
  orange = "#D96A1D",
  light  = "#E3E6EA",
  dark   = "#4B525A"
)

gcps_boe5 <- c(
  area1 = "#2F5FB3",
  area2 = "#007C91",
  area3 = "#5E8C31",
  area4 = "#6A4CC3",
  area5 = "#D96A1D"
)

gcps_ramps <- list(
  maroon = c("#DDC7C7", "#BA8C8C", "#944C4C", "#660000", "#540000"),
  blue   = c("#D1DCEE", "#A1B7DD", "#6D8FCA", "#2F5FB3", "#274E93"),
  teal   = c("#C7E2E7", "#8CC4CE", "#4CA3B2", "#007C91", "#006677"),
  green  = c("#DCE6D2", "#B7CBA2", "#8EAE6F", "#5E8C31", "#4D7328"),
  violet = c("#DED8F2", "#BCAEE4", "#9782D5", "#6A4CC3", "#573EA0"),
  orange = c("#F7DECD", "#EEBC99", "#E49761", "#D96A1D", "#B25718"),
  neutral = c("#F4F5F7", "#E3E6EA", "#B6BCC4", "#7A828C", "#4B525A")
)

gcps_diverging <- list(
  maroon_teal = c("#7A2E2E", "#C9A6A6", "#F3F4F6", "#9DCFD7", "#007C91"),
  blue_orange = c("#4268B8", "#B7C8EA", "#F3F4F6", "#F3C9A8", "#D96A1D")
)

gcps_pairs <- list(
  brand_vs_alt = c("#660000", "#007C91"),
  blue_orange  = c("#2F5FB3", "#D96A1D"),
  teal_orange  = c("#007C91", "#D96A1D"),
  violet_green = c("#6A4CC3", "#5E8C31"),
  maroon_blue  = c("#660000", "#2F5FB3")
)

# Helpers ------------------------------------------------------------------

text_color <- function(hex) {
  rgb <- grDevices::col2rgb(hex)
  luminance <- (0.299 * rgb[1, ] + 0.587 * rgb[2, ] + 0.114 * rgb[3, ]) / 255
  ifelse(luminance > 0.62, "#1F2328", "#FFFFFF")
}

swatch_row <- function(cols, labels = names(cols), height = 82) {
  if (is.null(labels)) labels <- rep("", length(cols))
  pieces <- mapply(
    FUN = function(col, lab) {
      fg <- text_color(col)
      sprintf(
        paste0(
          "<div style='flex:1; min-width:110px; background:%s; color:%s; border-radius:10px; ",
          "padding:12px; height:%spx; display:flex; flex-direction:column; justify-content:space-between; ",
          "box-shadow: inset 0 0 0 1px rgba(0,0,0,.08);'>",
          "<div style='font-weight:700; font-size:0.95rem;'>%s</div>",
          "<div style='font-family:monospace; font-size:0.9rem;'>%s</div>",
          "</div>"
        ),
        col, fg, height, lab, col
      )
    },
    cols,
    labels,
    SIMPLIFY = FALSE,
    USE.NAMES = FALSE
  )

  cat(
    "<div style='display:flex; gap:12px; flex-wrap:wrap; margin:0.5rem 0 1.2rem 0;'>",
    paste(pieces, collapse = ""),
    "</div>"
  )
}

rule_card <- function(title, use_when, typical, rule) {
  cat(sprintf(
    paste0(
      "<div style='border:1px solid #D9DDE3; border-radius:12px; padding:16px; margin:0.8rem 0 1rem 0; background:#FFFFFF;'>",
      "<div style='font-size:1.05rem; font-weight:700; margin-bottom:8px;'>%s</div>",
      "<p><strong>Use when:</strong> %s</p>",
      "<p><strong>Typical charts:</strong> %s</p>",
      "<p style='margin-bottom:0;'><strong>Rule:</strong> %s</p>",
      "</div>"
    ),
    title, use_when, typical, rule
  ))
}

mini_bar <- function(cols, heights = NULL) {
  if (is.null(heights)) heights <- c(52, 68, 84, 102, 118)[seq_along(cols)]
  pieces <- mapply(
    FUN = function(col, h) sprintf(
      "<div style='width:42px; height:%spx; background:%s; border-radius:6px 6px 0 0;'></div>",
      h, col
    ),
    cols,
    heights,
    SIMPLIFY = FALSE,
    USE.NAMES = FALSE
  )
  cat(
    "<div style='display:flex; gap:8px; align-items:flex-end; padding:10px 0 2px 0;'>",
    paste(pieces, collapse = ""),
    "</div>"
  )
}

mini_diverging <- function(cols) {
  pieces <- vapply(
    cols,
    function(col) sprintf(
      "<div style='flex:1; height:28px; background:%s;'></div>",
      col
    ),
    character(1)
  )
  cat(
    "<div style='display:flex; overflow:hidden; border-radius:8px; border:1px solid #D9DDE3; margin:0.3rem 0 0.8rem 0;'>",
    paste(pieces, collapse = ""),
    "</div>"
  )
}
```

<style>
.callout-note {
  border-left-color: #660000;
}
.table-smaller table {
  font-size: 0.95rem;
}
</style>

## Purpose

This page defines a practical color system for GCPS analytics products. It is designed for:

- dashboard use on computer screens
- presentation use on projection screens
- public-facing summary pages
- consistent color decisions across Quarto, Shiny, Power BI, and static reports

The goal is not just to list hex codes. The goal is to specify **when to use which kind of scale**.

## Design logic

The system is organized into four parts:

1. **Brand anchor** for district identity
2. **Qualitative hues** for unordered categories such as Board Areas
3. **Sequential ramps** for ordered numeric values
4. **Diverging scales** for charts with a meaningful midpoint such as zero, district average, or target

::: {.callout-note}
## Operating rule
GCPS maroon should usually signal district identity, district totals, or a focal highlight. It should not behave like just another category unless that is intentional.
:::

## Base palette

### Brand + Board Areas + neutrals

```{r}
swatch_row(
  gcps_base,
  labels = c(
    "GCPS Maroon", "Blue", "Teal", "Green",
    "Violet", "Orange", "Light Neutral", "Dark Neutral"
  )
)
```

```{r}
gcps_base
```

### Board Area palette

```{r}
swatch_row(
  gcps_boe5,
  labels = c("Area 1", "Area 2", "Area 3", "Area 4", "Area 5")
)
```

```{r}
gcps_boe5
```

## Use cases

### 1. Brand / district identity

```{r}
rule_card(
  "Brand / district identity",
  "the chart needs to show GCPS overall, district totals, or the official district value",
  "KPI cards, district average reference lines, featured bars, district summary marks",
  "Use maroon as the anchor. Keep it scarce enough that it still signals importance."
)
```

```{r}
swatch_row(c("#660000"), labels = "GCPS")
```

### 2. Unordered category comparison

```{r}
rule_card(
  "Unordered category comparison",
  "the categories are different but not ranked, such as Board Areas or program types",
  "grouped bars, dot plots, multi-series lines, legends with category labels",
  "Use the qualitative palette. Do not imply order with unrelated categories."
)
mini_bar(gcps_boe5, heights = c(78, 96, 64, 88, 72))
swatch_row(gcps_boe5, labels = c("Blue", "Teal", "Green", "Violet", "Orange"), height = 72)
```

### 3. Ordered low-to-high values

```{r}
rule_card(
  "Ordered low-to-high values",
  "the data move in one direction from low to high, such as rate, count, or share",
  "heatmaps, choropleths, shaded tables, ranked bars with intensity",
  "Use a single-hue sequential ramp. Darker should mean more."
)
mini_bar(gcps_ramps$blue)
swatch_row(gcps_ramps$blue, labels = paste("Step", 1:5), height = 70)
```

### 4. Below / at / above a meaningful midpoint

```{r}
rule_card(
  "Centered around a meaningful midpoint",
  "the midpoint represents zero change, district average, target, or another meaningful reference value",
  "difference heatmaps, centered bar charts, performance vs target maps",
  "Use a diverging scale only when the midpoint actually matters."
)
mini_diverging(gcps_diverging$maroon_teal)
swatch_row(gcps_diverging$maroon_teal, labels = c("Low", "Below", "Center", "Above", "High"), height = 64)
```

### 5. Highlight one thing against many others

```{r}
rule_card(
  "Highlight one value against many others",
  "one district, one school, or one selected entity needs to stand out from a background set",
  "ranked bars, lollipops, strip plots, comparison tables",
  "Use one focal color and keep all others neutral."
)
mini_bar(c("#B6BCC4", "#B6BCC4", "#660000", "#B6BCC4", "#B6BCC4"), heights = c(62, 78, 104, 85, 70))
swatch_row(c(focal = "#660000", others = "#B6BCC4"), labels = c("Focal", "Muted peers"), height = 72)
```

### 6. Within-one-group composition

```{r}
rule_card(
  "Within-one-group composition",
  "the categories are parts of the same parent concept, such as sub-statuses within one population",
  "stacked bars, waffles, treemaps, segmented cards",
  "Use a monochrome family so the viewer reads the pieces as related."
)
mini_bar(gcps_ramps$teal, heights = c(28, 46, 66, 88, 112))
swatch_row(gcps_ramps$teal, labels = paste("Step", 1:5), height = 70)
```

### 7. Ordered status or severity

```{r}
rule_card(
  "Ordered status or severity",
  "the categories have a natural progression such as low, medium, high or approaching, meeting, exceeding",
  "summary matrices, scorecards, simple heat tables",
  "Use a light-to-dark ordered family rather than unrelated hues."
)
swatch_row(c(low = "#DCE6D2", medium = "#8EAE6F", high = "#5E8C31"), labels = c("Low", "Medium", "High"), height = 72)
```

### 8. Current year versus prior year

```{r}
rule_card(
  "Two-time-point comparison",
  "the story is current versus prior year or before versus after",
  "paired bars, slope charts, dumbbells, two-line trends",
  "Make the current value darker and more saturated than the prior value."
)
swatch_row(c(current = "#660000", prior = "#B6BCC4"), labels = c("Current", "Prior"), height = 72)
```

### 9. Actual, target, and benchmark

```{r}
rule_card(
  "Actual, target, and benchmark",
  "one measured value needs to be shown against a target or reference",
  "bullet charts, annotated KPI cards, line charts with reference bands",
  "Keep the actual saturated, keep the target neutral, and keep the benchmark band subtle."
)
swatch_row(c(actual = "#007C91", target = "#4B525A", band = "#E3E6EA"), labels = c("Actual", "Target", "Benchmark band"), height = 72)
```

### 10. Heatmap or matrix values

```{r}
rule_card(
  "Heatmap / matrix values",
  "a table or grid is encoding a single numeric variable",
  "school-by-year matrices, subgroup-by-metric tables, month-by-rate grids",
  "Use a sequential ramp and keep the legend obvious."
)
swatch_row(c("#F4F8F9", "#C7E2E7", "#8CC4CE", "#4CA3B2", "#007C91"), labels = paste("Bin", 1:5), height = 68)
```

### 11. Geographic choropleth

```{r}
rule_card(
  "Map fills",
  "a geographic unit is colored by a single numeric variable or by deviation from a benchmark",
  "choropleths and zone maps",
  "Use sequential for one-way magnitude and diverging for deviation from a meaningful center."
)
swatch_row(c("#F7DECD", "#EEBC99", "#E49761", "#D96A1D", "#B25718"), labels = paste("Seq", 1:5), height = 68)
mini_diverging(c("#7A2E2E", "#D7B4B4", "#F3F4F6", "#A8D2DA", "#007C91"))
```

### 12. Background, scaffolding, and UI support

```{r}
rule_card(
  "Neutrals for interface and scaffolding",
  "the purpose is support, not emphasis",
  "background panels, striping, borders, gridlines, labels, muted bars",
  "Most of the interface should be neutral so the data colors can do the work."
)
swatch_row(c(light = "#E3E6EA", mid = "#B6BCC4", dark = "#4B525A"), labels = c("Light", "Mid", "Dark"), height = 72)
```

## Sequential ramps

### Maroon
```{r}
swatch_row(gcps_ramps$maroon, labels = paste("Step", 1:5), height = 72)
```

### Blue
```{r}
swatch_row(gcps_ramps$blue, labels = paste("Step", 1:5), height = 72)
```

### Teal
```{r}
swatch_row(gcps_ramps$teal, labels = paste("Step", 1:5), height = 72)
```

### Green
```{r}
swatch_row(gcps_ramps$green, labels = paste("Step", 1:5), height = 72)
```

### Violet
```{r}
swatch_row(gcps_ramps$violet, labels = paste("Step", 1:5), height = 72)
```

### Orange
```{r}
swatch_row(gcps_ramps$orange, labels = paste("Step", 1:5), height = 72)
```

### Neutral
```{r}
swatch_row(gcps_ramps$neutral, labels = paste("Step", 1:5), height = 72)
```

## Diverging scales

### Maroon to teal
```{r}
swatch_row(gcps_diverging$maroon_teal, labels = c("Low", "Below", "Center", "Above", "High"), height = 72)
```

### Blue to orange
```{r}
swatch_row(gcps_diverging$blue_orange, labels = c("Low", "Below", "Center", "Above", "High"), height = 72)
```

## Recommended pairs

```{r}
for (nm in names(gcps_pairs)) {
  cat(sprintf("<h3>%s</h3>", gsub("_", " ", nm)))
  swatch_row(gcps_pairs[[nm]], labels = c("A", "B"), height = 68)
}
```

## Do / don't rules

::: {.callout-note}
## Do

- Use **maroon** for GCPS identity, district totals, and focal highlights.
- Use **five qualitative hues** for Board Areas and other unordered categories.
- Use **one hue** for ordered numeric scales.
- Use **diverging color** only when the center has meaning.
- Use **neutrals** heavily in the interface so comparisons stay clear.
:::

::: {.callout-note}
## Don't

- Do not use five unrelated Board Area colors for a continuous choropleth.
- Do not use maroon as one of many equal colors unless that is intentional.
- Do not mix too many dark colors in one chart for projection-heavy use.
- Do not use diverging colors just because they look interesting.
- Do not let background elements compete with the data marks.
:::

## Implementation objects

```{r}
gcps_colors <- list(
  brand = list(
    gcps = "#660000"
  ),
  qualitative = list(
    boe5 = gcps_boe5
  ),
  sequential = gcps_ramps,
  diverging = gcps_diverging,
  neutral = list(
    light = "#E3E6EA",
    mid   = "#B6BCC4",
    dark  = "#4B525A"
  )
)

gcps_colors
```

## ggplot2 helpers

```{r}
scale_color_gcps_qual <- function(...) {
  ggplot2::scale_color_manual(values = gcps_boe5, ...)
}

scale_fill_gcps_qual <- function(...) {
  ggplot2::scale_fill_manual(values = gcps_boe5, ...)
}

scale_fill_gcps_ramp <- function(
  palette = c("maroon", "blue", "teal", "green", "violet", "orange", "neutral"),
  ...
) {
  palette <- match.arg(palette)
  ggplot2::scale_fill_manual(values = gcps_ramps[[palette]], ...)
}

scale_color_gcps_ramp <- function(
  palette = c("maroon", "blue", "teal", "green", "violet", "orange", "neutral"),
  ...
) {
  palette <- match.arg(palette)
  ggplot2::scale_color_manual(values = gcps_ramps[[palette]], ...)
}
```